MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-14 docs/paper/out/paper.tex
fix references to figures -- paper

The MG7 analysis workflow is summarized in Figure X. The input files for
MG7 are the FASTQ files resulting from a paired-end NGS sequencing
experiment.

@@

All the reads are assigned under two different algorithms of assignment:
i. Lowest Common Ancestor based taxonomic assignment (LCA) and ii. Best
BLAST Hit based taxonomic assignment (BBH). Figure X displays
schematically the LCA algorithm applied sensu stricto (left panel) and
the called `in line' exception (right panel) designed in order to gain
specificity in the assignments in the cases in which the topology of the
taxonomical nodes corresponding to the BLAST hits support sufficiently
the assignment to the most specific taxon.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity (by default \(evalue \leq e^{-15}\))
filtering those hits that are not sufficiently good comparing them with
the best one. We select the best HSP (High Similarity Pair) per
reference sequence and then choose the best HSP (that with lowest e
value) between all the selected ones. The bitscore of this best HSP
(called S) is used as reference to filter the rest of HSPs. All the HSPs
with bitscore below p x S are filtered. p is a coefficient fixed by the
user to define the bitscore required, e.g.~if p=0.9 and S=700 the
required bitscore threshold would be 630. Once we have the definitive
HSPs selected, we obtain their corresponding taxonomic nodes using the
taxonomic assignments that NCBI provides for all the nt database
sequences. Now we have to analyze the topological distribution of these
nodes in the taxonomy tree: i. If all the nodes forms a line in the
taxonomy tree (are located in a not branched lineage to the tree root)
we should choose the most specific taxID as the final assignment for
that read. We call to this kind of assignment the `in line' exception
(see Figure X right panel). ii. If not, we should search for the
\emph{sensu stricto} Lowest Common Ancestor (LCA) of all the selected
taxonomic nodes (See Figure X left panel). In this approach we decided
to use the bitscore for evaluating the similarity because it is a value
that increases when similarity is higher and depends a lot on the length
of the HSP. Some reads could not find sequences with enough similarity
in the database and then they would be classified as reads with no hits.
Advanced metagenomics analysis approaches \citep{huson2013microbial}
have adopted LCA assignment algorithms because it provides fine and
trusted taxonomical assignment.

>>>>

The MG7 analysis workflow is summarized in Figure 1. The input files for
MG7 are the FASTQ files resulting from a paired-end NGS sequencing
experiment.

@@

All the reads are assigned under two different algorithms of assignment:
i. Lowest Common Ancestor based taxonomic assignment (LCA) and ii. Best
BLAST Hit based taxonomic assignment (BBH). Figure 2 displays
schematically the LCA algorithm applied sensu stricto (left panel) and
the called `in line' exception (right panel) designed in order to gain
specificity in the assignments in the cases in which the topology of the
taxonomical nodes corresponding to the BLAST hits support sufficiently
the assignment to the most specific taxon.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity (by default \(evalue \leq e^{-15}\))
filtering those hits that are not sufficiently good comparing them with
the best one. We select the best HSP (High Similarity Pair) per
reference sequence and then choose the best HSP (that with lowest e
value) between all the selected ones. The bitscore of this best HSP
(called S) is used as reference to filter the rest of HSPs. All the HSPs
with bitscore below p x S are filtered. p is a coefficient fixed by the
user to define the bitscore required, e.g.~if p=0.9 and S=700 the
required bitscore threshold would be 630. Once we have the definitive
HSPs selected, we obtain their corresponding taxonomic nodes using the
taxonomic assignments that NCBI provides for all the nt database
sequences. Now we have to analyze the topological distribution of these
nodes in the taxonomy tree: i. If all the nodes forms a line in the
taxonomy tree (are located in a not branched lineage to the tree root)
we should choose the most specific taxID as the final assignment for
that read. We call to this kind of assignment the `in line' exception
(see Figure 2 right panel). ii. If not, we should search for the
\emph{sensu stricto} Lowest Common Ancestor (LCA) of all the selected
taxonomic nodes (See Figure 2 left panel). In this approach we decided
to use the bitscore for evaluating the similarity because it is a value
that increases when similarity is higher and depends a lot on the length
of the HSP. Some reads could not find sequences with enough similarity
in the database and then they would be classified as reads with no hits.
Advanced metagenomics analysis approaches \citep{huson2013microbial}
have adopted LCA assignment algorithms because it provides fine and
trusted taxonomical assignment.
